---
title: "SpatialGAMmod User Manual and Workflow"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{SpatialGAMmod User Manual and Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# SpatialGAMmod: User Manual and Workflow

This document provides a step-by-step workflow to analyze spatial transcriptomics data using the **SpatialGAMmod** package. It guides you through key functions for assigning spatial regions, modeling spatial gene expression gradients with GAMs, computing cell type proportions, comparing tumor vs non-tumor regions, and visualizing results.

---

## Prerequisites

- A Seurat object with spatial transcriptomics data, including:
  - Expression data (e.g., assay `"RNA"`, slot `"data"`)
  - Spatial coordinates in metadata columns (default: `"imagecol"` and `"imagerow"`)
  - Cell type annotations in metadata (e.g., `"CellType"`)
  
- The SpatialGAMmod package installed (see README for installation).

- Required R packages: `Seurat`, `mgcv`, `ggplot2`, `viridis`, `dplyr`, `tidyr`, `pheatmap`.

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE, results = FALSE)
```

## Assign Cells to Spatial Contour Regions
Use spatial coordinates or clustering to assign cells into spatial regions (contours) that represent biologically meaningful zones such as tumor core, invasive front, stroma, etc.

```{r}
library(SpatialGAMmod)

# Assuming seurat_obj is pre-loaded
seurat_obj <- assign_contour_regions(
  seurat_obj,
  coord_cols = c("imagecol", "imagerow"),
  num_contours = 5,
  method = "kmeans"
)

# This adds a metadata column (default name "assigned_region") to seurat_obj
head(seurat_obj@meta.data$assigned_region)
```

## Calculate Cell Type Proportions per Region
Compute the proportions of each cell type within each spatial region.


```{r}
prop_df <- calculate_celltype_proportions(
  seurat_obj,
  region_col = "assigned_region",
  celltype_col = "CellType"
)

head(prop_df)
```

## Visualize Cell Type Proportions as Heatmaps
Visualize these proportions across regions using heatmaps.

```{r}
proportion_heatmaps(
  prop_df,
  region_col = "assigned_region",
  celltype_col = "CellType",
  proportion_col = "proportion",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  scale = FALSE
)
```

## Fit Spatial GAM Gradients for Genes or Signatures
Model spatial gradients of gene expression or gene signatures using Generalized Additive Models (GAMs). This step smooths expression over spatial coordinates and optionally fits separate models per group (e.g., tumor cores).
```{r}
genes_to_model <- c("ACTA2", "COL1A1")
signature_list <- list(
  my_signature = c("ACTA2", "COL1A1", "FAP")
)

gam_res <- spatial_gam_gradient(
  seurat_obj,
  genes = genes_to_model,
  signature_list = signature_list,
  split_by = "tumor_core_location",  # optional grouping
  spatial_coords = c("imagecol", "imagerow"),
  plot_results = TRUE
)

# View GAM fit summaries
print(gam_res$metadata)

# Plot smoothed spatial expression gradients (example for first plot)
print(gam_res$plots[[1]])
```

## Predict GAM Expression on a Spatial Grid
Generate smooth spatial predictions for visualization or further analysis:

```{r}
# Choose a GAM model from the list returned by spatial_gam_gradient()
gam_model <- gam_res$gam_fits[[1]]

# Predict expression on a spatial grid
grid_pred <- predict_gam_to_grid(
  gam_model,
  seurat_obj@meta.data,
  coord_cols = c("imagecol", "imagerow"),
  grid_length = 100
)

# Visualize grid predictions
library(ggplot2)
ggplot(grid_pred, aes(x = imagecol, y = imagerow, fill = predicted)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed() +
  labs(title = "Predicted Spatial Expression") +
  theme_minimal()
```

## Compare Tumour vs Non-Tumour Regions
Perform differential expression analysis between tumor and non-tumor regions defined by contour assignments.

```{r}
diff_res <- tumour_vs_non_tumour_contour(
  seurat_obj,
  contour_col = "assigned_region",
  tumour_regions = c("tumour_core"),
  non_tumour_regions = c("stroma", "immune"),
  genes = genes_to_model,
  plot_results = TRUE
)

# Inspect results
head(diff_res$results)

# Plot volcano plot
print(diff_res$plot)
```

## Additional Analyses

### Spatial Autocorrelation (Moran’s I)

```{r}
moran_i <- spatial_autocorrelation(
  seurat_obj,
  gene = "ACTA2",
  coord_cols = c("imagecol", "imagerow"),
  neighbors_k = 6
)
print(moran_i)
```

### Spatial Correlation Decay with Distance

```{r}
dist_corr <- spatial_correlation_with_distance(
  seurat_obj,
  gene = "ACTA2",
  coord_cols = c("imagecol", "imagerow"),
  n_bins = 20,
  sample_size = 10000
)
plot(dist_corr$mean_distance, dist_corr$mean_correlation, type = "b",
     xlab = "Distance", ylab = "Expression similarity",
     main = "Spatial correlation decay with distance")
```

### Run Pseudotime Within Regions

```{r}
seurat_obj <- run_pseudotime_within_region(
  seurat_obj,
  region_col = "assigned_region",
  reduction = "pca",
  dims = 1:10,
  cluster_col = "CellType"
)
```